---
title: "Conferência de inventários florestais em campo"
subtitle:
titlerunning: Conferência 10%
date: "`r format(Sys.time(), '%d %b %Y %H:%M:%S %Z')`"
author: "Eric B. Gorgens, UFVJM"
output:
  bookdown::word_document2: default
  pdf_document: default
  bookdown::html_document2: default
  bookdown::pdf_book:
    base_format: rticles::springer_article
    extra_dependencies: booktabs
abstract: "A mock computed manuscript created in RStudio using {Rmarkdown}. The {Bookdown}
  and {Rticles} packages were used to output the text in Springer Nature's desired
  manuscript format. \n"
bibliography: bibliografia.bib
biblio-style: spbasic
authors:
- name: José Jhones Matuda
  address: Universidade Federal dos Vales do Jequitinhonha e Mucuri, Campus JK, Diamantina, MG, Brasil. CEP 39100-000
  email: conferir@ufvjm.edu.br
- name: Tiago Dornelas Otini
  address: Universidade Federal dos Vales do Jequitinhonha e Mucuri, Campus JK, Diamantina, MG, Brasil. CEP 39100-000
  email: conferir@ufvjm.edu.br
- name: Marcio Romarco Leles de Oliveira
  address: Universidade Federal dos Vales do Jequitinhonha e Mucuri, Campus JK, Diamantina, MG, Brasil. CEP 39100-000
  email: conferir@ufvjm.edu.br
- name: Gilciano Saraiva Nogueira
  address: Universidade Federal dos Vales do Jequitinhonha e Mucuri, Campus JK, Diamantina, MG, Brasil. CEP 39100-000
  email: conferir@ufvjm.edu.br
- name: Eric Bastos Gorgens
  address: Universidade Federal dos Vales do Jequitinhonha e Mucuri, Campus JK, Diamantina, MG, Brasil. CEP 39100-000
  email: eric.gorgens@ufvjm.edu.br
csl: nature.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      echo = FALSE)
```

```{r load-libraries, include=FALSE}
# load libraries
library(tidyverse)
library(ggbeeswarm)
library(bookdown)
```

# Introdução {#intro}

"Literate programming" is a style of programming that uses computational notebooks to weave together code, explanatory text, data and results into a single document, enhancing scientific communication and computational reproducibility.[@shen2014; @perkel2018a; @perkel2018] (These references were added into the document using RStudio's integration with the open-source Zotero reference manager [@perkel2020] plus the [Better BibTeX](https://retorque.re/zotero-better-bibtex/) Zotero plugin.)

Several platforms for creating such documents exist.[@perkel2021] Typically, these documents interleave code and text 'blocks' to build a computational narrative. But some, including [R Markdown](https://rmarkdown.rstudio.com/), [Observable](https://www.observablehq.com), and the [Jupyter Book](https://jupyterbook.org/intro.html) extension to the Jupyter ecosystem, also allow authors to include and execute code "inline" -- that is, within the text itself.

This makes it possible to create fully executable manuscripts in which the document itself computes and inserts values and figures into the text rather than requiring authors to input them manually. This is in many ways the 'killer feature' of computed manuscripts: it circumvents the possibility that the author will enter an incorrect number, or forget to update a figure or value should new data arise. Among other uses, that allows authors to automatically time-stamp their documents, or insert the current version number of the software they use into their methods. For instance, this document was built at **`r format(Sys.time(), "%d %b %Y %H:%M:%S %Z")`** and calls the following R packages: `{tidyverse}` ver. **`r packageVersion("tidyverse")`**, `{ggbeeswarm}` ver. **`r packageVersion("ggbeeswarm")`** and `{bookdown}` ver. **`r packageVersion("bookdown")`**.

In this manuscript, created in RStudio using the R Markdown language, we will demonstrate a more practical example. (An Observable version is [also available](https://observablehq.com/@jperkel/example-executable-observable-notebook).)

# Material e métodos {#material}


# Resultados {#resultados}

Imagine we are analyzing data from a clinical trial. We have grouped subjects in three bins and measured the concentration of some metabolite. (These data are simulated.)

```{r initial-data}
# read in some initial data
#df1 <- read_csv('data/example-data-1.csv')
```

```{r radius}
# radius of a circle
r = 10
```

Rather than analyzing those data and then copying the results into our manuscript, we can use the programming language `R` to do that in the manuscript itself. Simply enclose the code inside backticks, with the letter `r`. For instance, we could calculate the circumference and area of a circle `r r`:

$$A = \pi r^2, C = 2 \pi r$$

# Discussão {#discussao}

You could write "A = `` `r
pi * r^2` `` and C = `` `r

# Conclusão {#conclusao}

2 * pi * r` ``". Plugging in the radius *r* = **`r r`**, that evaluates to "A = **`r round(pi * r^2, 2)`** and C = **`r round(2 * pi * r, 2)`**".

# References